Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-05-26 paper/msg.tex
Edits on the FE section, and on the smoother section.

FIXME MCD = multilevel constraint decomposition, a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020}

@@

We may now define the ice dynamics operator (IDO), a map using the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
                        \bzero, & \text{otherwise}, \end{cases} \label{eq:ido}
\end{equation}
where $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, thus permitting substantial irregularities.

@@

Suppose $\mathcal{T}$ is a triangulation of $\Omega$, which we assume is polygonal.  (If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on their low expected regularity at the ice margin, we will represent the surface elevations $s\in \mathcal{K}$ using the piecewise-linear ($P_1$) finite element (FE) space $\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega)$.  Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.

Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$, and define two closed and convex subsets of $\mathcal{V}^h$:
\begin{equation}
\mathcal{K}^h = \{r \in \mathcal{V}^h \,:\, r \ge b^h \text{ and } r|_{\partial\Omega} = b^h|_{\partial\Omega}\}, \qquad \mathcal{D}_+^h = \{t \in \mathcal{V}^h \,:\, t \ge 0\}. \label{eq:feKD}
\end{equation}
Let $(x_j,y_j)$ denote the coordinates an arbitrary node of the mesh $\mathcal{T}$, and $\psi_j \in \mathcal{V}^h$ the corresponding hat function, i.e.~such that $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.

With these FE choices, the discrete SIGP computes the FE approximation $s^h \in \mathcal{K}^h$ satisfying
\begin{equation}
F^h(s^h)[r - s^h] \ge 0 \quad \text{for all } r \in \mathcal{K}^h , \label{eq:fesigpweakform}
\end{equation}
where $F^h(s^h)[r] = - \int_\Omega (\Phi^h(s^h) + a)\, r \,dx dy$; compare \eqref{eq:sigpfunctional}.  Here $\Phi^h$ is an FE approximation of the IDO $\Phi$; see below.  We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.

All FE computations in this paper use the Python library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.  For $s^h \in \mathcal{K}^h$ we will compute $\Phi^h(s^h)$, namely the numerical Glen-Stokes \eqref{eq:glenstokesweak} surface velocity associated to the surface elevation $s^h$, using a extruded-mesh approximation of $\Lambda_s^h \subset \RR^{d+1}$, the icy domain as defined by \eqref{eq:lambdas}.  Specifically, if $d=2$ and if $\mathcal{T}$ is a triangulation of $\Omega \subset \RR^d$, and given $s^h$ and $b^h$ from $\mathcal{V}^h$, we apply the Firedrake concept of an extruded mesh \cite{McRaeetal2016} to give a prism element decomposition of $\Lambda_s^h$.  If $d=1$ then the extruded mesh has quadrilateral elements; a sketch is shown in Figure \ref{fig:extruded}.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
\end{center}
\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh.  In the ``pinched'' extrusion (left), $s^h$ defines where the trace of $\bu^h$ is evaluated (strong).  In the ``cliffs'' extrusion (right) all elements have a minimum thickness $H_{\text{min}} / m_z$, so the trace evaluation may be above the original $s^h$ (dotted).}
\label{fig:extruded}
\end{figure}

FIXME computing $\Phi^h(s)$

@@

As a first step toward solving \eqref{eq:fesigpweakform} we propose (slow) solvers which will be smoothers in our multilevel method (next section)  The smoothers are the projected nonlinear Gauss-Seidel and projected nonlinear Jacobi (PNJ) methods \cite{KinderlehrerStampacchia1980}.  In fact we will describe both smoothers but only implement PNJ.

Given a current surface elevation $s\in \mathcal{K}^h$, a smoother will sweep through the nodes in a fixed order, solving one-dimensional VI problems, as follows, and updating $s$.   Note that, because $\psi_p \in \mathcal{D}_+^h$, if $r \in \mathcal{K}^h$ then $r + c \psi_p \in \mathcal{K}^h$ for all $c\ge b(x_p,y_p) - r(x_p,y_p)$.  Therefore let $\omega_p^s = b(x_p,y_p) - s(x_p,y_p)$.  The one-dimensional VI at $p$ implied by \eqref{eq:fesigpweakform} is to find $c \ge \omega_p^s$ such that
\begin{equation}
F^h(s+c\psi_p)[(s+\gamma \psi_p) - (s+c \psi_p)] \ge 0 \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \omega_p^s$.  Defining $F_p^h(c) = F^h(s+c\psi_p)[\psi_p]$, \eqref{eq:fepointwiseviEARLY} is equivalent to the simplified form
\begin{equation}
(\gamma - c) F_p^h(c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
Supposing \eqref{eq:fepointwisevi} is well-posed and solved for $c$, the surface elevation $s$ is then updated $s \gets s + c \psi_p$ by the smoother.  (Observe that admissibility is preserved: the updated surface elevation is also in $\mathcal{K}^h$.)  Each pointwise update is done immediately by the (serial) PNGS smoother, while the (parallel) PNJ smoother computes solutions to \eqref{eq:fepointwisevi} at all nodes $p$ before the updates.

If $F_p^h(c)$ were linear and increasing then the solution of \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_p^s\right\}$ if $F_p^h(c) = \alpha c + \beta$ with $\alpha > 0$, but in fact $F_p^h(c)$ is nonlinear.  The simplest approximation scheme is thus to linearize $F_p^h(c)$, that is, smooth using one projected Newton step for the one-dimensional VI problem at each node.  To implement this strategy, note that \eqref{eq:sigpfunctionalderiv} and \eqref{eq:idoderiv} imply
\begin{equation}
F_p^h(c) = F^h(s+c\psi_p)[\psi_p] \approx F^h(s)[\psi_p] + c\, (F^h)'(s,\psi_p)[\psi_p] = \alpha c + \beta \label{eq:fepointwisefunctionalapprox}
\end{equation}
where
\begin{align}
\alpha &= \int_\Omega \left(- \left(\lim_{\eps\to 0^+} \frac{\bu^h|_{s+\eps \psi_p} - \bu^h|_s}{\eps}\right) \cdot \bn_s + \Phi^h(s) \cdot \left<(\psi_p)_x,(\psi_p)_y,0\right>\right)\, \psi_p \,dx dy \label{eq:fepointwisealpha} \\
\beta &= - \int_\Omega (\Phi^h(s) + a)\, \psi_p \,dx dy \label{eq:fepointwisebeta}
\end{align}
Each smoother application will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak} on the geometry determined by the current surface elevation.  In fact, supposing we have computed $\bu^h|_s$ and $\Phi^h(s)$ then several terms in these formulas are straightforward FE computations over the support of $\psi_p$.

>>>>

FIXME MCD = multilevel constraint decomposition, a multigrid \cite{Trottenbergetal2001} method basically by \cite{Tai2003} for obstacle problems; see \cite{Bueler2022};  obstacle problem view first extended to Stokes by \cite{WirbelJarosch2020};  The computations in this paper use the Python FE library Firedrake \cite{Rathgeberetal2016}, which applies an embedded domain language \cite{Alnaesetal2014} to convert weak forms into discrete equations which are solved in parallel using the PETSc \cite{Balayetal2020} library.

@@

We may now define the ice dynamics operator (IDO), a map using the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}, \end{cases} \label{eq:ido}
\end{equation}
where $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, thus permitting substantial irregularities.

@@

Assume $\Omega \subset \RR^d$ is polygonal and suppose $\mathcal{T}$ is a triangulation of $\Omega$.  (If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
\begin{equation}
\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega).
\end{equation}
Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.

Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$.  From $b^h$ we define the closed and convex admissible subset of $\mathcal{V}^h$:
\begin{equation}
\mathcal{K}^h = \{r^h \in \mathcal{V}^h \,:\, r^h \ge b^h \text{ and } r^h|_{\partial\Omega} = b^h|_{\partial\Omega}\}.  \label{eq:feK}
\end{equation}
For $s^h\in \mathcal{K}^h$ we also define the numerical icy domain as in \eqref{eq:lambdas}:
\begin{equation}
\Lambda_{s^h} = \{(x,y,z)\,:\,b^h(x,y) < z < s^h(x,y)\} \subset \RR^{d+1}.  \label{eq:felambdas}
\end{equation}
Also, a triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in the $s^h$ geometry if $s^h=b^h$ at every vertex of $T$, and icy otherwise.

Given $s^h$ and $b^h$ from $\mathcal{V}^h$, we construct a Firedrake extruded mesh \cite{McRaeetal2016} to decompose $\Lambda_{s^h}$.  If $d=2$ then the extruded mesh consists of triangular prisms, while if $d=1$ it has quadrilateral elements.  The extruded mesh construction starts from the triangulation $\mathcal{T}$ of $\Omega \subset \RR^d$, now called the ``base mesh''.  As shown in Figure \ref{fig:extruded}, each icy $T$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.  However, ice-free $T$ have no extruded mesh elements at all; these columns are empty.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
\end{center}
\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.  In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold); see \eqref{eq:ido}.  In the ``cliffs'' extrusion (right) all elements have a minimum thickness and the trace evaluation (bold) may occur above $s^h$ (dotted).}
\label{fig:extruded}
\end{figure}

We will compare two extrusion modes.  In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly, prism (quadilateral) elements at the ice margin degenerate; such elements have empty facets.  In the alternate ``cliffs'' extrusion, each element over an icy $T$ is nondegenerate.  Specifically, if $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in experiments, then each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$.  In this extrusion, a stress-free condition is applied on the cliffs  in the Glen-Stokes problem \eqref{eq:glenstokesweak}; they are treated the same way as the top surface.

Let $\Phi^h:\mathcal{K}^h \to (\mathcal{V}^h)'$ be the FE approximation of the IDO $\Phi$.  By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:glenstokesweak} on the domain $\Lambda_{s^h}$.  That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, but on $\overline{\partial} \Lambda_{s^h}$, and then extend by zero to all of $\Omega$.  In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.

We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}, and a Newton iteration, with direct solution of the step equations and back-tracking line search, to resolve the power-law nonlinearity.  As stated in Section \ref{sec:intro}, note that our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme (next section) are needed to solve the SIGP \eqref{eq:sigpweakform}.  Overall optimality of the solution algorithm would also need an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.

Finally, the discrete SIGP computes $s^h \in \mathcal{K}^h$ satisfying
\begin{equation}
F^h(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \quad \text{for all } r^h \in \mathcal{K}^h , \label{eq:fesigpweakform}
\end{equation}
where $F^h(s)[r] = - \ip{\Phi^h(s)}{r}$; compare \eqref{eq:sigpfunctional}.  We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.

@@

In this section we propose solvers for \eqref{eq:fesigpweakform}.  While they are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.  We will describe both projected nonlinear Gauss-Seidel and Jacobi methods \cite{KinderlehrerStampacchia1980}, but only implement the latter.

To describe these methods suppose $(x_i,y_i)$ denotes an arbitrary vertex of $\mathcal{T}$, and $\psi_i \in \mathcal{V}^h$ the corresponding hat function, i.e.~such that $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.  Also let
\begin{equation}
\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}.  \label{eq:feDplus}
\end{equation}
Note that, because $\psi_i \in \mathcal{D}_+^h$, if $r^h \in \mathcal{K}^h$ then $r + c \psi_i \in \mathcal{K}^h$ for all $c\ge \omega_i^s$ where $\omega_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.

Given a current surface elevation $s^h\in \mathcal{K}^h$, a smoother will sweep through the nodes in a fixed order, solving one-dimensional VI problems and then updating $s^h$.  The one-dimensional VI at node $i$, derived from \eqref{eq:fesigpweakform}, finds $c \ge \omega_i^{s^h}$ such that
\begin{equation}
F^h(s^h+c\psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \omega_i^{s^h}$.  Let
\begin{equation}
\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i}, \label{eq:ferhoi}
\end{equation}
a point-wise residual function.  Then \eqref{eq:fepointwiseviEARLY} is equivalent to the simplified form
\begin{equation}
(\gamma - c) \,\rho_i(s^h; c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
Supposing \eqref{eq:fepointwisevi} is well-posed and solved for $c$, the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.  Observe that admissibility is preserved: the updated surface elevation is also in $\mathcal{K}^h$.

Each pointwise update is done immediately by the (serial) Gauss-Seidel smoother, while the (parallel) Jacobi smoother computes solutions to \eqref{eq:fepointwisevi} at all nodes $i$ before then doing all the updates.  In linear problems it is known that Gauss-Seidel generally converges in about half as many iterations \cite{Greenbaum1997}, but here there is a more glaring difference between the two iterations.  Gauss-Seidel would be an extraordinarily expensive approach because a new Glen-Stokes solution would be required at every node of $\mathcal{T}$.  We will prefer the Jacobi strategy; see below.

FIXME If $F_p^h(c)$ were linear and increasing then the solution of \eqref{eq:fepointwisevi} would be given by a simple formula \cite{GraeserKornhuber2009}, namely $c = \max\left\{-\beta/\alpha, \omega_p^s\right\}$ if $F_p^h(c) = \alpha c + \beta$ with $\alpha > 0$, but in fact $F_p^h(c)$ is nonlinear.  The simplest approximation scheme is thus to linearize $F_p^h(c)$, that is, smooth using one projected Newton step for the one-dimensional VI problem at each node.  To implement this strategy, note that \eqref{eq:sigpfunctionalderiv} and \eqref{eq:idoderiv} imply
\begin{equation}
F_p^h(c) = F^h(s+c\psi_p)[\psi_p] \approx F^h(s)[\psi_p] + c\, (F^h)'(s,\psi_p)[\psi_p] = \alpha c + \beta \label{eq:fepointwisefunctionalapprox}
\end{equation}
where
\begin{align}
\alpha &= \int_\Omega \left(- \left(\lim_{\eps\to 0^+} \frac{\bu^h|_{s+\eps \psi_p} - \bu^h|_s}{\eps}\right) \cdot \bn_s + \Phi^h(s) \cdot \left<(\psi_p)_x,(\psi_p)_y,0\right>\right)\, \psi_p \,dx dy \label{eq:fepointwisealpha} \\
\beta &= - \int_\Omega (\Phi^h(s) + a)\, \psi_p \,dx dy \label{eq:fepointwisebeta}
\end{align}
Each smoother application will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak} on the geometry determined by the current surface elevation.  In fact, supposing we have computed $\bu^h|_s$ and $\Phi^h(s)$ then several terms in these formulas are straightforward FE computations over the support of $\psi_p$.
